---
title: "Estimating ancestry proportions from allele frequencies only"
author: "Martha Richards"
output:
  prettydoc::html_pretty:
    theme: Cayman
    toc: true
    toc_depth: 3
---

<button id="back-to-top" onclick="window.scrollTo({top:0,behavior:'smooth'})" title="Back to top">↑</button>

```{css, echo=FALSE}
#back-to-top {
  position: fixed;
  bottom: 2rem;
  right: 2rem;
  background: #4a90d9;
  color: white;
  border: none;
  border-radius: 50%;
  width: 42px;
  height: 42px;
  font-size: 1.2rem;
  cursor: pointer;
  opacity: 0;
  transition: opacity 0.3s;
  z-index: 9999;
}
#back-to-top.visible { opacity: 1; }
```
```{js, echo=FALSE}
window.addEventListener('scroll', function() {
  var btn = document.getElementById('back-to-top');
  btn.classList.toggle('visible', window.scrollY > 300);
});
```
```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "C:/Users/c23043538/OneDrive - Cardiff University/Desktop/Ancestry_Inference")
```

```{r, include=FALSE}
setwd("C:/Users/c23043538/OneDrive - Cardiff University/Desktop/Ancestry_Inference")
```

# __Estimating Ancestry Proportions from Allele Frequencies only__

__What are Ancestry Proportions?__

Ancestry proportions or admixture proportions describe the ratio of an individual's genome that can be traced back to different ancestral populations. The estimations are important in human population genetics to understand an individual’s demographic history and to correct for population structure. Population structure needs to be corrected because genetic differences across groups can look like disease associations which can lead to false results. One way of doing this is through Principal Component Analysis, which is done in the bigsnpr work package. This asks instead of “does Variant X correlate with this trait,” but “does Variant X correlate with the trait after we have already accounted for the person’s position on the PCA map?” 

Ancestry proportions can give the percentage of where the individual is from geographically, for example, a person might have an estimated ancestry proportion of 60% West African, 30% European, and 10% East Asian. These are not family history estimates — they are calculated directly from patterns in an individual's DNA. The results from this analysis for each individual is a set of proportions that sum to 1.
How they are calculated:
They can be calculated using a range of tools but here we explore two different methods from the bigsnpr package.

__Why are Ancestry Proportions important?__

Ancestry Proportions are important in many aspects’ bioinformatics. In population genetics, they reveal historical patterns of migration admixture between human populations which helps to reconstruct how modern populations formed in the past. 

Ancestry inference is important in clinical medicine, as disease prevalence and drug responses often vary across ancestral backgrounds. This variation occurs not because genes change location, but because the frequency of specific risk variants and their genetic correlations (linkage disequilibrium) differ globally. For example,  a genetic variant which increased your chance of having schizophrenia might be very common (found in 40% of people) in population A, but the same allele might be very rare (1%) in population B. The gene is in the same place for both groups, but the “risk” is more frequent in population A.

These variant alleles are usually inherited with their neighbour genes called linkage disequilibrium. In one ancestry a harmless “tag” variant might be inherited alongside a hidden harmful variant, yet in another ancestry, thousands of years might have broken that linkage. So, if a researcher looks for that “tag” neighbour allele in the second ancestry, they might think the patient is at risk when they weren’t. The signal has changed because the neighbour alleles are different.

Incorporating these ancestry proportions into Polygenic Risk Scores (PRS) significantly improves their predictive accuracy, leading to better treatment outcomes for all patients. This is particularly crucial for addressing health inequities, as genetic datasets have historically been Eurocentric; accounting for diverse ancestry proportions helps identify and bridge the gaps in our current genomic knowledge.


### __Preparation__

__Load required R libraries__

```{r, results='hide', warning=FALSE, message=FALSE}
library(dplyr)
library(quadprog)
library(bigsnpr)
library(DT)
library(bigstatsr)
library(prettydoc)
```

```{r, include=FALSE}
setwd("C:/Users/c23043538/OneDrive - Cardiff University/Desktop/Ancestry_Inference")
```

__Download the Data used in the Tutorials__

```{r, eval=FALSE}
"https://figshare.com/ndownloader/files/38019072", 
"https://figshare.com/ndownloader/files/38019027",  
"https://figshare.com/ndownloader/files/38019024", 
"https://figshare.com/ndownloader/files/38077323",
"https://figshare.com/ndownloader/files/38247288", 
"http://www.epigad.org/gwas_ilae2018_16loci/all_epilepsy_METAL.gz"
"https://figshare.com/ndownloader/files/31620968"
"https://figshare.com/ndownloader/files/31620953"
bigsnpr::download_plink("tmp-data")   
bigsnpr::download_plink2("tmp-data")  
```

__Set up file pathway to name and store data__

```{r}
DIR <- "tmp-data/datasets"
```

This line stores a file path as a text string by creating a variable *DIR* This does not create the folder, access it or manipulate it, it just stores the path as a text. This makes coding easier because you don't need to re-write the whole line which helps with reproducibility, because if you ever move your data to a different folder, you only need to change this one line. 

__Naming and Storing Files__

```{r, results='hide'}
gz <- runonce::download_file(
  "http://www.epigad.org/gwas_ilae2018_16loci/all_epilepsy_METAL.gz",
  dir = "../datasets")
readLines(gz, n = 3)
```

* This creates a variable called _gz_ which will store the file path once it has been downloaded from the internet using the runonce code. The key feature is it only downloaded once, if the file already exists it skips the download, saving time and computational power.
* *dir = ".../datasets"* tells R where to save the downloaded file on the computer. 
* *ReadLines* looks at the first three lines of the file to check the right page has been downloaded.

__Read in summary statistics, keeping only 5 key columns__

```{r, results='hide'}
sumstats <- bigreadr::fread2(
  gz, select = c("CHR", "BP", "Allele2", "Allele1", "Freq1"), 
  col.names = c("chr", "pos", "a0", "a1", "freq") 
) %>%
  mutate_at(3:4, toupper) 
```

* This creates a variable called *sumsats* which stores the data once it has been read in using the *fread2* function from the bigreadr package which can read large files quickly.
* The *select* function only reads these 5 columns from the file, ignoring everything else. 
* The *col.names* renames those 5 columns into shorter cleaner names. 
* The _mutate_at_ function from the dplyr package applies a transformation to specific columns, and mutate means to change or modify. This code takes columns 3 and 4 (a0 and a1) and converts any lowercase letters to UPPERCASE, because alleles should always be uppercase (A, T, C, G) and the source file may have mixed cases

```{r, results='hide'}

all_freq <- bigreadr::fread2(
  runonce::download_file("https://figshare.com/ndownloader/files/31620968",
                         dir = DIR, fname = "ref_freqs.csv.gz"))
projection <- bigreadr::fread2(
  runonce::download_file("https://figshare.com/ndownloader/files/31620953",
                         dir = DIR, fname = "projection.csv.gz"))
```

* This code is essentially naming files.
* This creates a variable called all_freq which will store the data once it has been coded. 
* The _bigreadr::_ specifies that the _fread2_ function comes from the bigsnpr package. It is a reading function which reads really large files, like genetic data. 
* The file at the end of the function is the path to the file being read, csv means it is a spreadsheet style with columns separated by commas.

### __Analysis__

__Match the epilepsy GWAS variants to the reference panel by chromosome, position and alleles__

We need to match variants within the summary statistics with the ones from the reference by chromosome, position and alleles.

This is important for analysis because the two data sets are not automatically compatible with each other. 

The *sumstats* from the epilepsy GWAS data contains;

* Variants found in epilepsy patients

While the all_freq (reference panel) contains: 

* Variants from known ancestry reference populations

```{r}
matched <- snp_match(
  mutate(sumstats, chr = as.integer(chr), beta = 1),
  all_freq[1:5]
) %>%
  mutate(freq = ifelse(beta < 0, 1 - freq, freq))
```

This code takes the summary statistics and the reference frequency data, finds all the variants which appear in both data sets and makes sure the allele frequencies are correct after matching.

* *snp_match* is a function from bigsnpr which matches SNPs between two data sets. It matches by chromosome, position and alleles to find the same variants.
* Before matching the _mutate_ makes two small changes to _sumstats_. _chr = as.integer(chr)_ converts the chromosome from a text to a whole number because _snp_match()_ requires it to be a number not text (e.g. "1" to 1). _beta = 1_ adds a new column called beta and sets every values to 1, which is a placeholder value required by _snp_match_ to work.
* *all_freq[1:5]* is the reference data set being matched against and only the first 5 columns are being used because they contain the chromosome, position and allele information needed for matching.
* *mutate(freq = ifelse(beta < 0, 1 - freq, freq))* corrects the allele frequency after matching. 
* *ifelse(beta < 0, 1 - freq, freq)* means: if beta is negative → replace freq with 1 - freq, f beta is positive → keep freq as it is. This is important because when _snp_match_ flips an allele to make it match, the frequency also needs to be flipped, since all frequencies must add up to 1, if one allele is 0.3 the other must be 0.7.

__Correct allele frequencies for any flipped alleles after matching__

```{r}
correction <- c(1, 1, 1, 1.008, 1.021, 1.034, 1.052, 1.074, 1.099,
                1.123, 1.15, 1.195, 1.256, 1.321, 1.382, 1.443)
```

* *correction* creates a variable of 16 numbers, which are the pre-calculated correction factors for each principal component. These values were calculated by the *bigsnpr* package. 
* This is needed because when you project a new data set onto a principal component space which was built from a different reference data set there is mathematical shrinkage effect. This step rules out systematical bias.

__Estimate ancestry proportions using the snp_ancestry_summary function__

```{r}
res <- snp_ancestry_summary(
  freq = matched$freq,
  info_freq_ref = all_freq[matched$`_NUM_ID_`, -(1:5)],
  projection = projection[matched$`_NUM_ID_`, -(1:5)],
  correction = correction)
```

Takes the allele frequencies and estimates what proportion of the sample comes from each ancestry group.

* *snp_ancestry_summary()* is the main function from bigsnpr which actually estimates ancestry proportions. This takes the allele frequencies and estimates what proportion of the sample comes from each ancestry group.
* *freq = matched$freq* This takes the freq column from the matched dataframe. These are the allele frequencies observed in the epilepsy patients and is the dataset you are trying to analyse.
* *info_freq_ref = all_freq[matchedNUM_ID, -(1:5)]*,  provides the reference population allele frequencies to compare against. 
* *matched$_NUM_ID* selects only the rows that were successfully matched earlier
* *projection = projection[matched$NUM_ID, -(1:5)]* provides the PC projection matrix for the match variants. This is used to project your data into the same PC space as the reference populations.

__Apply correction factors to remove mathematical shrinkage bias__

* *correction = correction)* - Inside the same block, which passes in the 16 pre-calculated correction numbers

__Merge similar regional groups together__

This next code takes the raw ancestry results and merges groups within similar regions together into broader categories. Next the proportions are added up within each group and printed as rounded percentages. 

This is important because the reference panel has detailed ancestry groups which for most research is too refined. Merging them into broader groups makes the results easier to interpret, more statistically reliable and more comparable.


```{r}
group <- colnames(all_freq)[-(1:5)]
group[group %in% c("Scandinavia", "United Kingdom", "Ireland")]   <- "Europe (North West)"
group[group %in% c("Europe (South East)", "Europe (North East)")] <- "Europe (East)"
grp_fct <- factor(group, unique(group))

final_res <- tapply(res, grp_fct, sum)
round(100 * final_res, 1)
```

* *colnames()* gets all the column names from all_freq. _[-(1:5)]_ removes the first 5 columns which are the chromosome, position and allele information. This leaves the ancestry groups names which were used as column headers in the reference panel. So _group is now a list of ancestry names like "Africa"
* %in%* means contained in. This finds any ancestry labels which are "Scandinavia", "United Kingdom" or "Ireland" and mergers them into one broader group called "Europe." This is then repeated with Eastern European groups.
* *grp_fct <- factor(group, unique(group))* converts the _group_ variable into a factor (Rs way of representing categories). R needs the groups to be factors to be able to sum them up correctly in the next step. 
* *unique(group)* ensures each group name only appears once. 
* *tapply()* is a function which takes the ancestry proportion results in _res_ and adds them up within each group. So for example all the individual European subgroup proportions get summed into one single European proportion. 
* The result is stored in _final_res_

__Convert proportions into percentages__

_round(100 * final_res, 1)_.  The times 100 converts the proportions from decimals to percentages, which are then rounded to 1 decimal place.

### __Results__

Europe (North West): 68.0% — the dominant ancestry
Europe (East): 13.9%
Finland: 6.5%
Italy: 3.4%
Asia (East): 3.1%
Ashkenazi: 1.8%

These results show the aggregated ancestry proportion estimates by combining sub-populations into broader regional groups. This data shows that North East Europe is the dominant ancestry signal at 68.0% followed by Eastern Europe (13.9%.)

This approach is important because you don't need individual-level reference genomes, just the population level allele frequencies (which was the input and also publicly available.) This makes the analysis more computationally lightweight. Furthermore, having the results in percentages makes the data more readable unlike PC-based methods which require additional interpretation.

__Limitations__

This analysis assumes the reference populations are representative from all groups and non-overlapping which is not true. An example of this is Finnish ancestry is partly shared with Eastern Europe but has been grouped with Western Europe. Moreover, the ~4% scattered across Africa/Asia/Japan likely reflects reference panel bleed — slight allele frequency overlaps between distant populations — rather than true ancestry. This method cannot detect _admixture timing_.

